\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-29 chapters/covert_align.tex
Layout chapter covert align except fig 1.4 and 1.5

\chapter{Compensating \acs{erp} jitter for gaze-independence}%
\label{sec:covert-align}
\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ \emph{A
Scandal in Bohemia -- Sir Arthur Conan Doyle}}
\noindent\emph{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different \ac{vsa} conditions (overt, covert, and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \textcite{Treder2010}, we presented six flashing targets (without letters or
symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}%
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip
	\begin{subfigure}{\linewidth}
		\caption{}%
		\label{fig:erps}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
	\end{subfigure}%

  \caption[Stimulation interfaces and evoked \acsp{erp}.]{%
		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split \ac{vsa} to the cued target.
		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split \ac{vsa}.
    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

@@

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_1.tikz.tex}
    \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
    \acs{bci}~\cite{Treder2010}.}%
    \label{fig:covert-align/interface/interface}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_2.tikz.tex}
    \caption[Intensification.]{When a target is intensified, it is enlarged for
    a very short time.}%
    \label{fig:covert-align/interface/intensification}
  \end{subfigure}

@@

\clearpage
\begin{subappendices}
  \section{Single-trial classification performances}
  \begin{table}[ht]
  \centering
  \makebox[\textwidth][c]{%
    \input{chapters/covert_align_performance.tex}
  }
  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
curve for all evaluated models, visuospatial attention conditions, and
  datasets.}
  \label{tab:covert-align/results/single-trial}
\end{table}

  \clearpage
  \section{Cross-condition transfer classification performances}

  \begin{table}[ht]
      \footnotesize
      \input{chapters/covert_align_cross-CVSA-ERP.tex}
      \caption[Transfer learning results for CVSA-ERP]{Cross-condition transfer
      classification performance (cross-validated \ac{rocauc})
      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
      on one \ac{vsa} condition and tested across all
      \ac{vsa}
      conditions to evaluate gaze-independence and transfer learning
      performance.}
  \end{table}
  \begin{table}[ht]
      \footnotesize
      \input{chapters/covert_align_cross-BNCI2014-009.tex}
      \caption[Transfer learning results for BNCI2014-009]{Cross-condition transfer
      classification performance (cross-validated \ac{rocauc})
      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
      on one \ac{vsa} condition and tested across all \ac{vsa}
      conditions to evaluate gaze-independence and transfer learning
      performance.}
  \end{table}
\end{subappendices}

>>>>

\chapter{Compensating \acs{erp} jitter for gaze-independence}%
\label{sec:covert-align}
\epigraph{``You see, but you do not observe!''}{Sherlock Holmes in  \\ \emph{A
Scandal in Bohemia -- Sir Arthur Conan Doyle}}
\publicationnote{This chapter was published as part of~\textcite{VanDenKerchove2024}.}

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
consists of 15 participants, mean age 26.34$\pm$3.04 years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different \ac{vsa} conditions (overt, covert, and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \textcite{Treder2010}, we presented six flashing targets (without letters or
symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}[t]
		\includegraphics[width=\linewidth]{figures/covert_align/dataset_vsa.pdf}
    \caption[Dataset stimulation interfaces.]{%
    Interfaces and visuospatial attention (\ac{vsa})
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split \ac{vsa} to the cued target.
		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.

    }%
		\label{fig:interface}
\end{figure*}
\begin{figure*}[t]
  \input{figures/covert_align/grand_avg.pfg}
	%\includegraphics[width=\linewidth]{figures/covert_align/figure3b.pdf}
  \caption[Evoked \acs{erp} per dataset and condition.]{%
		Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split \ac{vsa}.
    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
		\label{fig:erps}
\end{figure*}

@@

\begin{figure}[t]
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_1.tikz.tex}
    \caption[The stimulation interface.]{The stimulation interface based on the visual Hex-o-Spell
    \acs{bci}~\cite{Treder2010}.}%
    \label{fig:covert-align/interface/interface}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_2.tikz.tex}
    \caption[Intensification.]{When a target is intensified, it is enlarged for
    a very short time.}%
    \label{fig:covert-align/interface/intensification}
  \end{subfigure}

@@

\clearpage
\begin{subappendices}
  \section{Single-trial classification performances}
  \begin{table}[ht]
  \input{chapters/covert_align_performance.tex}
    \caption[Classification performances.]{Cross-validated single-trial
    classification \ac{rocauc} (\%) for all evaluated models, visuospatial attention conditions, and
    datasets.}
    \label{tab:covert-align/results/single-trial}
  \end{table}
  \clearpage
  \section{Cross-condition transfer classification performances}
  \begin{table}[ht]
      \input{chapters/covert_align_cross-CVSA-ERP.tex}
      \caption[Transfer learning results for CVSA-ERP]{Cross-condition transfer
      classification performance (cross-validated \ac{rocauc})
      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
      on one \ac{vsa} condition and tested across all
      \ac{vsa}
      conditions to evaluate gaze-independence and transfer learning
      performance.}
  \end{table}
  \clearpage
  \begin{table}[ht]
      \input{chapters/covert_align_cross-BNCI2014-009.tex}
      \caption[Transfer learning results for BNCI2014-009]{Cross-condition transfer
      classification performance (cross-validated \ac{rocauc})
      for all evaluated models in the CVSA-ERP dataset. Each decoder is trained
      on one \ac{vsa} condition and tested across all \ac{vsa}
      conditions to evaluate gaze-independence and transfer learning
      performance.}
  \end{table}
\end{subappendices}
